Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-06-27 hazelnut-popl17.tex
tweaks

\emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text. %Structure editors for professional programmers are also under active development (e.g. \texttt{mbeddr}.)

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful. 
Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner. To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.

We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself. To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut. We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application. Our reference implementation is written using \lstinline{js_of_ocaml}.

>>>>

\emph{Structure editors} allow programmers to edit the tree structure of a program directly. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text. %Structure editors for professional programmers are also under active development (e.g. \texttt{mbeddr}.)

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not only syntactically well-formed but also statically and dynamically meaningful.  
Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct terms in an unnatural ``outside-in'' manner. To avoid this problem, Hazelnut's {action semantics} automatically places terms assigned a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems.

We intend Hazelnut to serve as a principled foundational calculus, rather than as an end-user tool itself. To that end, we demonstrate how Hazelnut's rich metatheory, which we have mechanized in Agda, guides an extension to Hazelnut. We also  discuss how Hazelnut's semantics lends itself to implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml}.

### 2016-06-27 intro.tex
tweaks

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program.

Every day, programmers use text-based tools to construct and manipulate programs.  
These programs are written in languages which are defined by a textual syntax. 
While textual syntax has proved to have practical utility, it also introduces some fundamental complexity into the programming process.

%Doesn't really start out with a bang, IMHO. I like to start papers out with a "big" problem.
%Most programming languages define a textual syntax. This allows programmers to use text editors and other standard text-based tools to construct and manipulate programs. While this is of substantial practical utility, it also introduces some fundamental complexity into the programming process.

Cognitively, it requires that programmers understand the subtleties of text parsing (e.g. the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.
One study~\cite{Altadmri:2015:MCI:2676723.2677258} of novice programmer error found that the most common of all errors was an syntactic error, specifically unbalanced parenthesis.
% Too much detail?

It also complicates matters for tool designers because they might be confronted with text that does not correspond to a well-formed, meaningful program
By analyzing the same data set used in~\cite{6883030}, which contains 1,460 hours of fine-grained code editing logs collected from 21 programmers, we found that 44.2\% of the code edits resulted in syntactically malformed state of the source code.
%\todo{mention YoungSeok's data here}. 
This may be because the programmer is in the midst of a sequence of 
edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake. The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g. by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)

>>>>

%Although programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Most contemporary programming languages define a textual syntax so that programmers can use text editors and other text-based tools to construct and manipulate programs. While this design is of substantial practical utility, it also introduces some fundamental complexity into the programming process.

Cognitively, it requires that programmers understand the subtleties of text parsing (e.g. the relative precedence of the various forms.) This can be particularly challenging for novice and end-user programmers.\todo{citation?}

It also complicates matters for tool designers because tools are often confronted with text that does not correspond to a well-formed, meaningful program\todo{mention YoungSeok's data here}. This may be because the programmer is in the midst of a sequence of 
edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake. The language definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action (e.g. by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.)
